---
title: "Baby names"
author: "Karl Broman"
date: "`r Sys.Date()`"
output: html_document
---

```{r global_opts, include=FALSE}
knitr::opts_chunk$set(warning=FALSE, message=FALSE)
```

We first want to load some packages.

```{r load_packages}
library(babynames)
library(dplyr)
library(ggplot2)
library(DT)
library(plotly)
```

I'm interested in the relative use of the names "Karl" and "Carl". Let's filter out just those rows from the `babynames` data.

```{r, filter_karl}
karl <- babynames %>%
    filter(name=="Karl" | name=="Carl")
```

The results include `r nrow(karl)` rows.

Here's an interactive table, using the package [DT](https://github.com/rstudio/DT):

```{r datatable}
DT::datatable(karl)
```

Plot the results.

```{r lineplot}
ggplot(karl, aes(x=year, y=prop, color=name)) + geom_line()
```

That's weird. Turns out there were a small number of female Karls.

```{r lineplot_bysex}
ggplot(karl, aes(x=year, y=prop, color=name, linetype=sex)) +
    geom_line()
```

Here's an interactive version using the package [plotly](https://plot.ly/r/).

```{r plotly, fig.width=12, fig.height=7}
p <- ggplot(karl, aes(x=year, y=prop, color=name, linetype=sex)) + geom_line()
plotly::ggplotly(p)
```